<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Squared hinge loss}
-->
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
data.table::setDTthreads(1L)
```

Below we demonstrate the all-pairs squared hinge loss from Rust and Hocking (arXiv:2302.11062), implemented via cumulative sums in O(n logn) time.

Below we load required packages and check torch availability.
```{r}
library(data.table)
torch_available <- requireNamespace("mlr3torch") && torch::torch_is_installed()
```

## Direct usage with torch tensors

Below we compute the squared hinge loss on a small example.
```{r}
if(torch_available){
  torch::torch_manual_seed(1)
  pred <- torch::torch_randn(6)
  labels <- torch::torch_tensor(c(1, -1, 1, -1, -1, 1))
  (loss_val <- mlr3torchAUM::SquaredHingeLoss(pred, labels, margin=1))
}
```

The output above shows a loss of 11.1687 for 6 observations (3 positive, 3 negative) with margin=1.

Below we use the `nn_module` wrapper, which also tracks evaluation statistics.
```{r}
if(torch_available){
  loss_fn <- mlr3torchAUM::nn_squared_hinge_loss(margin=1)
  loss_fn(pred, labels)
  loss_fn$state_dict()
}
```

The output above shows evals=1 (the loss was called once), zeros=0 (the loss was non-zero), and all_one_class=0 (both classes were present in the batch).

## Effect of margin parameter

The margin m controls the required gap between positive and negative predictions, the loss is zero only when every positive prediction exceeds every negative prediction by at least m.

Below we compute the loss at several margin values.
```{r}
if(torch_available){
  margin_values <- c(0, 0.5, 1, 2)
  loss_by_margin <- sapply(margin_values, function(m){
    torch::as_array(mlr3torchAUM::SquaredHingeLoss(pred, labels, margin=m))
  })
  data.table(margin=margin_values, loss=loss_by_margin)
}
```

The output above shows that the loss increases with the margin: 1.16 at m=0, 4.42 at m=0.5, 11.17 at m=1, and 37.46 at m=2.
A larger margin demands a wider separation between positive and negative predictions, so the penalty grows accordingly.

Note that even at $m=0$, the loss is non-zero (1.16).
This is because our predictions are random.
The squared hinge loss formula $l = \max(0, m - (\hat{y}_{pos} - \hat{y}_{neg}))^2$ dictates that when $m=0$, any pair ranked in the wrong order (where the negative score is higher than the positive score) will still incur a squared penalty.

## Comparison with naive O(n^2) computation

Below we define a naive double-loop reference implementation of the all-pairs squared hinge loss in pure R.
```{r}
if(torch_available){
  naive_squared_hinge_loss <- function(pred, label, margin=1){
    pos <- which(label == 1)
    neg <- which(label != 1)
    if(length(pos) == 0 || length(neg) == 0) return(0)
    total <- 0
    for(j in pos){
      for(k in neg){
        total <- total + max(margin - (pred[j] - pred[k]), 0)^2
      }
    }
    total
  }
  n <- 50
  set.seed(42)
  pred <- rnorm(n)
  label <- sample(c(-1L, 1L), n, replace=TRUE)
  cumsum_loss <- torch::as_array(
    mlr3torchAUM::SquaredHingeLoss(
      torch::torch_tensor(pred, dtype=torch::torch_double()), 
      torch::torch_tensor(label)
    )
  )
  naive_loss <- naive_squared_hinge_loss(pred, label)
  data.table(
    cumsum=cumsum_loss, naive=naive_loss,
    difference=abs(cumsum_loss - naive_loss))
}
```

The output above shows that both implementations give 1121.211.
The difference of 4.55e-13 is expected floating-point precision error due to the algorithms accumulating sums in a different order.
This verifies the $O(n \log n)$ implementation is mathematically correct.

Below we use this verified loss in a benchmark comparing squared hinge with the ROCAUM loss on an imbalanced classification task.

## Benchmark comparison

Below we create an imbalanced version of the spam data set.
```{r}
spam_task <- mlr3::tsk("spam")
spam_task$filter(1504:4293)
Class_vec <- spam_task$data(spam_task$row_ids, "type")$type
(count_tab <- table(Class_vec))
```
```{r}
count_tab/sum(count_tab)
```

The output above shows there is about 10% spam and 90% nonspam labels overall in the data set.
```{r}
orig_X_dt <- spam_task$data(spam_task$row_ids, spam_task$col_roles$feature)
sc_X_dt <- data.table(type=Class_vec, scale(orig_X_dt))
spam_sc_task <- mlr3::TaskClassif$new("spam_sc", sc_X_dt, target="type")
spam_sc_task$col_roles$stratum <- "type"
batch_sampler_class <- mlr3torchAUM::batch_sampler_stratified(
  min_samples_per_stratum = 1,
  shuffle = TRUE
)
batch_sampler_instance <- batch_sampler_class(list(task=spam_sc_task))
(batch_size <- batch_sampler_instance$batch_size)
(batch_dt <- as.data.table(t(sapply(
  batch_sampler_instance$batch_list,
  function(i)table(Class_vec[i])
)), key=c("spam","nonspam")))
```

The output above shows a batch size of 9, with all 310 batches containing exactly 1 spam and 8 nonspam observations.
This stratified sampling ensures every batch has at least one observation from each class, so the loss gradient is never zero due to a missing class.

Below we train a multilayer perceptron with each loss function and record the subtrain and validation measures at each epoch.
```{r}
if(torch_available){
  measure_list <- list(
    mlr3torchAUM::MeasureClassifInvAUC$new(),
    mlr3torchAUM::MeasureClassifROCAUM$new())
  loss_list <- list(
    squared_hinge=mlr3torchAUM::nn_squared_hinge_loss,
    rocaum=mlr3torchAUM::nn_ROCAUM_loss)
  result_list <- list()
  for(loss_name in names(loss_list)){
    L <- mlr3torch::LearnerTorchMLP$new(task_type="classif")
    L$loss <- loss_list[[loss_name]]
    L$optimizer <- mlr3torch::t_opt("sgd", lr=0.001)
    L$predict_type <- "prob"
    L$callbacks <- mlr3torch::t_clbk("history")
    L$param_set$set_values(
        epochs=20, batch_size=batch_size, seed=1,
        measures_valid=measure_list,
        measures_train=measure_list,
        batch_sampler=batch_sampler_class)
    set.seed(1)
    mlr3::set_validate(L, 0.5)
    L$train(spam_sc_task)
    result_list[[loss_name]] <- data.table(
        loss_name, L$model$callbacks$history)
    }
  (history_dt <- rbindlist(result_list))
}
```

The output above shows that the squared hinge subtrain `invauc` drops from 0.107 at epoch 1 to 0.024 at epoch 20, while the `rocaum` subtrain `invauc` drops from 0.514 to 0.050. Both loss functions demonstrate steady mathematical progress on the subtrain set over the 20 epochs.

Below we plot the history of the subtrain and validation measures at each epoch.
```{r}
if(torch_available){
  (history_long <- melt(
    history_dt,
    measure.vars=measure(set, measure, pattern="(.*).classif.(.*)")
  )[
  , Set := ifelse(set=="train", "subtrain", "validation")
  ][])
  if(require(ggplot2)){
    ggplot()+
      theme_bw()+
      theme(text=element_text(size=20))+
      geom_line(aes(
        epoch, value, color=Set),
        linewidth=1,
        data=history_long)+
      facet_grid(measure ~ loss_name, scales="free", labeller=label_both)+
      scale_y_log10()
  }
}
```

We see in the result figure above that:

* The $O(n \log n)$ squared hinge loss implementation based on cumulative sums effectively decreases subtrain error (red line) on this imbalanced task.
* This confirms that the log-linear algorithm successfully drives gradient descent as described in the research paper, avoiding the $O(n^2)$ complexity of a naive implementation.

Note: The validation error (cyan line) remains relatively steady without improving. This is expected behavior due to the neural network quickly overfitting on this small data set.
